---
title: "Statistical Analysis Report"
author: "Robert Munch"
date: "October 29, 2018"
output:
  pdf_document: default
  html_document: default
---

##Statistical Analysis

#Snow Water Equivalency Data
Using data gathered from the National Resource Conservation Service (NRCS) Snow Telemetry stations, the average snow water equivalent (SWE) in millimeters for the year was gathered and plotted along with a linear line showing the overall trend.
Annual averages were used rather than monthly averages, because of the summer months when it does not snow causes the data to appear cluttered and incoherant.
```{r}
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(readxl)
library(ggfortify)
library(fpp2)
library(GGally)

PrecipData <- read_csv("~/R/Capstone Project/PrecipData.txt", 
                       comment = "#")

#Removing the rows that contain incomplete data since giving them zeros would affect the final data.
#Numbering the months rather than names, and placing all the month values into their own column.
#Renaming station id and station name to be without spaces.
#filtering to only years 1980 to 2017, the first and last years with complete years worth of data.
#arranging by year to appear chronological.
PrecipData <- PrecipData %>%
  na.omit(PrecipData) %>%
  rename("01" = Jan, "02" = Feb , "03" = Mar, "04" = Apr, "05" = May, "06" = Jun, "07" = Jul, "08" = Aug, "09" = Sep, "10" = Oct, "11" = Nov, "12" = Dec) %>%
  rename(Station.Id = "Station Id", Station.Name = "Station Name") %>% 
  gather("Month", "SWE.mm", 4:15) %>% 
  filter(Year >= 1980 & Year <= 2017) %>%
  arrange(Year)

#Aggregating SWE data by Year, rewuiring SWE and Year to be numerics.
#Renaming columns, as names are lost in aggregating process.
SWE.year.agg <- aggregate(as.numeric(PrecipData$SWE.mm), list(as.numeric(PrecipData$Year)), mean)
colnames(SWE.year.agg) <- c("Year", "SWE.mm.Avg")

#Aggregating SWE data by month and Year, allowing data to be aggregated across all stations.
PrecipData.combine.date <- PrecipData %>% 
  unite("Date", "Year", "Month", sep = "-") %>% 
  arrange(Date)
PrecipData.station.agg <- aggregate(as.numeric(PrecipData.combine.date$SWE.mm), list(PrecipData.combine.date$Date), mean)
colnames(PrecipData.station.agg) <- c("Date", "SWE.mm")
PrecipData.station.agg <- separate(PrecipData.station.agg, "Date", c("Year", "Month"), sep = "-")

ggplot(SWE.year.agg, aes(x = Year, y = SWE.mm.Avg)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, aes(colour = "red")) +
  labs(y = "Snow Water Equivalent Annual Average", title = "Colorado Average Snow Water Equivlent Over Time") +
  theme(legend.position = "none")
```

Linear model for the annual average SWE data across all available sites in Colorado, as well as residual data.
```{r}
lm(SWE.mm.Avg ~ Year, SWE.year.agg)
summary(lm(SWE.mm.Avg ~ Year, SWE.year.agg))
checkresiduals(lm(SWE.mm.Avg ~ Year, SWE.year.agg))
```

Because the year 1981 seems to be especially low, and the available data is less complete earlier in the dataset, I wanted to make sure the variation was not due to certain monitoring stations only used in earlier data collection that had a higher chance of collecting low data. 

Using only sites available in 1981, I averaged annual SWE data and compared it to the original values. 
```{r}
#Showing 1981 is not an outlier
PrecipData.1981 <- PrecipData %>% 
  filter(Year == 1981) %>% 
  arrange(Station.Id) %>% 
  distinct(Station.Id) %>% 
  as.vector()

PrecipData.outlier <- PrecipData %>% 
  filter(Station.Id %in% unlist(PrecipData.1981))

PrecipData.outlier <- aggregate(as.numeric(PrecipData.outlier$SWE.mm), list(as.numeric(PrecipData.outlier$Year)), FUN = mean)  
colnames(PrecipData.outlier) <- c("Year", "SWE.mm.Avg.outlier")

PrecipData.outlier <- PrecipData.outlier %>%   
  bind_cols(SWE.year.agg) %>% 
  select(Year, SWE.mm.Avg.outlier, SWE.mm.Avg)

ggplot(PrecipData.outlier, aes(x = Year)) +
  geom_point(aes(y = SWE.mm.Avg.outlier, colour = "red")) +
  geom_smooth(method = "lm", se = FALSE, aes(y = SWE.mm.Avg.outlier, colour = "red")) +
  geom_point(aes(y = SWE.mm.Avg, colour = "blue")) +
  geom_smooth(method = "lm", se = FALSE, aes(y = SWE.mm.Avg, colour = "blue")) +
  labs(x = "Year", y = "Average annual SWE (mm)") +
  scale_colour_discrete(name = "", labels=c("Treatment", "Control"))
```
```{r}
chisq.test(PrecipData.outlier$SWE.mm.Avg.outlier, PrecipData.outlier$SWE.mm.Avg)
```
Based on the chart it is obvious the plot with only sites available in 1981 is very similar to data using all available sites.
Also, based on the p value from the chi-squared test, using  standard alpha value f 0.05, we cannot confirm the null hypothesis. Therefore the data do show that the low SWE value in 1981 is not due to a lack of stations.

#Atmospheric Carbon Dioxide concentration.
Using Data gathered from NOAA Earth System Research Laboratory, average monthly $CO_{2}$ concentration in ppm is displayed over time. Becasue the line is very linear, no linear model was generated. 
```{r}
#Importing atmospheric CO2 data from NOAA.
noaa.co2 <- read_table2("~/R/Capstone Project/noaa.co2.txt", 
                        col_names = FALSE, comment = "#")

#Naming columns that will be used.
#Removing data from outside the SWE data timeline.
noaa.co2 <- noaa.co2 %>% 
  rename(Year = X1, Month = X2, Interpolated.ppm = X5) %>% 
  filter(Year >= 1980 & Year <= 2017)

#Combine PrecipData.station.agg and noaa.co2 as a way to compare co2 ppm and SWE 
combine.data <- bind_cols(PrecipData.station.agg, noaa.co2) %>% 
  select("Year", "Month", "SWE.mm", "Interpolated.ppm")

#aggregting average co2 ppm by year
co2.year.agg <- aggregate(noaa.co2$Interpolated.ppm, list(noaa.co2$Year), mean)
colnames(co2.year.agg) <- c("Year", "co2.ppm.Avg")

ggplot(co2.year.agg, aes(x = Year, y = co2.ppm.Avg)) +
  geom_line() +
  labs(y = "Carbon Dioxide Concentration (ppm)", title = "Annual Atmospheric Carbon Dioxide Concentration (ppm)")
```
Because the direct data is so linear, no linear trend line was included in this model.
```{r}
SWE.co2.year.agg <- SWE.year.agg %>%
  bind_cols(co2.year.agg) %>%
  select(c(1,2,4))

lm.year.avg <- lm(SWE.mm.Avg ~ co2.ppm.Avg, data = SWE.co2.year.agg)
```
This plot shows the correlation between carbon ldioxide concentration and SWE values.
```{r}
ggplot(SWE.co2.year.agg, aes(x = co2.ppm.Avg, y = SWE.mm.Avg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(color = "red")) +
  labs(title = "Average Annual SWE (mm) and Carbon Dioxide Concentration (ppm)", x = "Average annual atmospheric CO2 (ppm)", y = "Average annual SWE (mm)") +
  theme(legend.position="none")
summary(lm.year.avg)
checkresiduals(lm.year.avg)
```

#Visitor Data
Because there was no publicly available data I could find for only the state of Colorado or to specific ski resorts, I had to use data made available by the National Ski Area Association, which gives visitor numbers by region. I decided to use the Rocky Mountain region. Colorado is in the Rocky Mountain region and visitor data averall should coorelate to Colorado quite fairly since Colorado one of the largest if not the largest state for ski visitors in the Rocky Mountain region.
Visitor data is used to see if there is a correlation with visitors to SWE data. Thus showing if lower snowfall causes a decrease in visitors to ski areas.
```{r}
ski.visits <- read.csv("~/R/Capstone Project/Ski visit data.csv")

#Renaming columns
#Becasue column name was changed from Visitor(millions), value must be changed as well.
#Removing years outside data window.
colnames(ski.visits) <- c("Year", "Visitors")
ski.visits <- ski.visits %>%
  mutate(Visitors = Visitors * 1e06) %>% 
  arrange(Year) %>% 
  filter(Year >= 1980)

#Combining visitors into data table with CO2 and SWE data. 
combine.all <-cbind(SWE.co2.year.agg, ski.visits$Visitors)
colnames(combine.all) <- c("Year", "SWE.mm", "CO2.ppm", "Visitors")

ggplot(ski.visits, aes(x = Year, y = Visitors)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, aes(color = "red")) +
  labs(title = "Visitors per year in Rocky Mountain Region") +
  theme(legend.position = "none")
```
Because the viditors will continulously be going up, and there are many other factors that influence why more people seem to be skiing each year, I plotted the residuals from the linear model and compared those to the residuals from the SWE model to compare if lower than average snowfall results in lower that average visitors. 
```{r}
lm.visitors <- lm(Visitors ~ Year, combine.all)
res.lm.visitors <- residuals(lm.visitors)

lm.SWE.year <- lm(SWE.mm ~ Year, combine.all)
res.lm.SWE.year <- residuals(lm.SWE.year)

ts.SWE.res <- ts(res.lm.SWE.year, start = 1980)
ts.visit.res <- ts((res.lm.visitors / 1e05), start = 1980)
SWE.Visit.res <- cbind(ts.SWE.res, ts.visit.res)
```
Plot showing correlation between residuals of SWE and Visitor data to show correlation of lower snow levels than average with lower number or visitors than average. 
```{r}
ggplot(as.data.frame(SWE.Visit.res), aes(x = ts.SWE.res, y = ts.visit.res)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(color = "red")) +
  labs(title = "Residuals from linear model of SWE and Visitor Data (1980 - 2017)",
       x = "SWE Residuals",
       y = "Visitor Residuals") +
  theme(legend.position = "none")

summary(SWE.Visit.res)
```
As shown on the plots, there is a positive correlation between the residuals for the Visitor data and the annual SWE data.

#Tourism
Tourism data was gathered from the Colorado department of tourism, and the thourism numbers represent total tourism dollars spent in the Mountain Reort area of Colorado.
```{r}
CO.tourism <- read_excel("CO tourism finacial data.xlsx")

#Changing tourism money to dollars
#Arranging Chronologically
CO.tourism <- CO.tourism %>% 
  mutate(Spending = Spending * 1e07) %>% 
  arrange(Year)

ggplot(CO.tourism, aes(x = Year, y = Spending)) +
  geom_line()+
  geom_smooth(method = "lm", se = FALSE, aes(color = "red")) +
  labs(title = "Tourism Spending in the Mountain Resort Area of Colorado") +
  theme(legend.position = "none")
```
Comparisons of all data correlations.
```{r}
#Because tourism spending data is only from 1996, other data must be restricted to include data from 1996 to 2017.  
combine.all.tourism <- combine.all %>% 
  arrange(Year) %>% 
  filter(Year >= 1996)
combine.all.tourism <- cbind(combine.all.tourism, CO.tourism$Spending)
colnames(combine.all.tourism) <- c("Year", "SWE.mm", "CO2.ppm", "Visitors", "Spending")
```
```{r}
ggpairs(combine.all.tourism[,2:5])
```
Residual comparisons for money spent and SWE.
```{r}
res.all <- 
  ts(cbind(
        residuals(lm(SWE.mm ~ Year, combine.all.tourism)),
        residuals(lm(CO2.ppm ~ Year, combine.all.tourism)),
        residuals(lm(Visitors ~ Year, combine.all.tourism)),
        residuals(lm(Spending ~ Year, combine.all.tourism))),
          start = 1996)
colnames(res.all) <- c("SWE.Residuals", "CO2.Residuals", "Visitors.Residuals", "Spending.Residuals")

ggplot(as.data.frame(res.all), aes(x = Spending.Residuals, y = SWE.Residuals)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(color = "red")) +
  labs(title = "Residuals of Spending and SWE", x = "Tourism Spending Residuals", y = "SWE Residuals") +
  theme(legend.position = "none")

summary(lm(SWE.Residuals ~ Spending.Residuals, res.all))
```
Comparisons of all residuals. 
```{r}
ggpairs(as.data.frame(res.all[,1:4]))
```
